Continous time Markov chain models; \\ extension, simulation and visualisation

### 2015-01-20 Response.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%---SUGGESTED SIMULATION SCENARIOS---%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Simulations}

\begin{itemize}

\subsection*{David Posada}

\item {
{\it
\textbf{
Simulate a number of epochs that do not correspond with the assumed ones.
}% END: textbf
}% END: it
}% END: item
\\
Running.

\item {
{\it
\textbf{
Simulating under a non-epoch model, for example where different process can apply to contemporaneous branches, would be fundamental to understand the robustness of the model.
}% END: textbf
}% END: it
}% END: item
\\
TODO.
% PL: I really think he suggest something like simulating with branch-specific omegas that on average increase from root to tips - infer with 2/3 epochs and check if we pick up increase
% FB: TODO. Need to implement.

\item {
{\it
\textbf{
Here it would have been interesting to see the correspondence between simulated and estimated tMRCAs when the true model has one epoch with omega=0.5 (or even better, omega equal to the weighted mean of the n-epoch model).
}% END: textbf
}% END: it
}% END: item
\\
Running.

@@

% 1
\item {
{\it
\textbf{
The thesis starts with a very nice introduction, which I think is very insightful for the reader who is not familiar with phylogenetic and/or Bayesian inference. 
In spite of this, I believe that the readability of the later chapters would be further improved if the introduction:
}% END: textbf
}% END: it
}% END: item

\begin{itemize}

% 1a
\item {
{\it
\textbf{
Briefly described the evolutionary models, such as GY94, that are later used.
}% END: textbf
}% END: it
}% END: item
\\
Done, there is now a \emph{Markov models of codon evolution} section in the introductory chapter.

% 1b
\item {
{\it
\textbf{
Briefly described how to accommodate time-stamped sequences, which is now considered in the analyses in Chapter 2, but not really discussed.
}% END: textbf
}% END: it
}% END: item
\\
Done, see Subsection~\ref{sub:dated_trees}.

% 1c
\item {
{\it
\textbf{
Give a bit more detail on phylogeography, e.g. how do the models look like and what is the connection between phylogeography and epoch models (some connection is described in Chapter 2, but remains a bit vague).
}% END: textbf
}% END: it
}% END: item
\\
FB: I don't understand this comment, we have a whole subsection on phylogeography (see Subsection~\ref{sub:phylogeo}).

### 2015-01-20 chapters/chapter01.tex
adding simulations
addressed comment on the dated tree tips

%GB: I think you should discuss some properties of EIDs at the nucleotide/amino acid/codon level (since that's what we've ended up extending) and link this to what you've done.
%Perhaps you can get some inspiration from the following paper: http://jvi.asm.org/content/84/19/9864.long
%Jeff Thorne is incorporating ''disease properties'' (for example, if they know that a certain non-synonymous mutation can be lethal) into his new codon models, explicitly taking into account likely lethal mutations.

%GB: Can you give a (or some) know example(s) of such substitution or the importance of some particular substitutions?
%How about the fact that a very dangerous virus (which made the news and hence readers will be able to relate to) is only a few mutations away from current bird flu viruses?
%Check this article (the links to the actual Nature and Science papers are at the bottom of the page): http://www.medicalnewstoday.com/articles/246964.php

@@

>>>>

@@

\subsection{Dated phylogenetic trees\label{sub:dated_trees}}
%how to accommodate time-stamped sequences

Time-stamped sequence data, which frequently arises from studies involving MEP populations for which sequences are sampled at different points in time and the evolutionary rates are high enough for significant changes to occur, can be used to calibrate the molecular clock and convert branch lengths into actual time units.

Fo simplicity let us assume a single site, where character sequence data was sampled on dates ranging from $1983$ to $2000$, as seen in Figure~\ref{fig:timing_sequences}.
We also assume that the tree topology is known and we want to estimate the unknown times $T_{0},T_{6},T_{7},T_{8}$.

\begin{figure}[H]
\centering
\includegraphics[scale=0.4]{timing_sequences} 
\caption{
{ \footnotesize 
{\bf Dated phylogenetic tree.} 
Discrete characters $CCACT$ observed at the external nodes (the tips of the tree) were sampled at different points in time.
The tree topology is known and assumed to be true.
The unknown dates of the internal nodes are labelled $T_{0},T_{6},T_{7},T_{8}$.
}% END: footnotesize
}
\label{fig:timing_sequences}
\end{figure}

We start by estimating the rate of mutation $r$, which here is assumed to apply to the whole tree (a strict clock), from the differences in sampling times and the genetic distance, given by the model of sequence substitution.
This allows us to change the scale of the tree from units of substitutions to a time-scale measured in years.
The likelihood of the tree given estimated parameters $r,T_{0},T_{6},T_{7},T_{8}$ can be calculated using methods presented in Subsection~\ref{sub:likelihood}.
